\section{Introduction}
\label{sec:introduction}
A notorious difficulty with intensional type theories like Martin-Löf
type theory or the calculus of inductive constructions is the lack of
extensionality principles in the core theory, and notably in its notion
of propositional equality. This makes the theory both inflexible with
respect to the notion of equality one might want to use on a given type
and also departs from the traditional equality notion from set
theory. Functional extensionality (pointwise equal functions are equal)
and propositional extensionality (logically equivalent propositions are
equal) are principles that are valid extensions of type theory and can
be internalized in an intensional theory, in the manner of Altenkirch
\textit{et al.}~\cite{altenkirch-mcbride-wierstra:ott-now}.  Another
extensionality principle coming from homotopy theory is univalence,
whose first instance is the idea that isomorphic types are
equal~\cite{Voevodsky:2011yq,Pelayo:2012uq}. All these principles should
be imposeable on a type theory like MLTT because, intuitively, no term
can distinguish between isomorphic types, pointwise equal functions or
logically equivalent propositions.
%
This hints at the idea that there ought to exist an internal model of
type theory where equality is defined on a type by type basis using
the aforementioned principles and a translation that witnesses that
any term of the theory is equipped with proofs that they respect these
properties. Formalizing a simple version of this definitional
translation is our first goal.

% However, before even starting to think about formalizing the
% translation, we need to enhance the universe system of \Coq to handle
% polymorphism. As the translation will be given as a set of definitions
% inside the assistant, we will have to instantiate these definitions at
% various levels depending on the source term. With fixed universes for
% each translated construct, we would quickly run into inconsistencies. We
% hence also present the design and implementation of a universe polymorphic
% extension of \Coq, that is conservative over the original system.

The central change in the theory is in the treatment of equality.  Much
interest has been devoted to the study of the identity type of type
theory and models thereof, starting with the groupoid model of Hofmann
and Streicher \cite{groupoid-interp}. This eventually led to the
introduction of homotopy type theory and the study of the ω-groupoid
model of type theory with identity types, which validates extensionality
principles.
% \cite{DBLP:journals/corr/abs-0812-0409}. 
This model in turn
guides work to redesign type theory itself to profit from its
richness, and develop a new theory that internalizes the new
principles. Preliminary attempts have been made, notably by Licata and
Harper \cite{DBLP:conf/popl/LicataH12} who develop a 2-dimensional
version of an hybrid intensional/extensional type theory which
integrates functional extensionality and univalence in the definition
of equality. Work is underway to make it more intensional, and this
starts by making the step to higher dimensions, whether finite (weak
n-groupoids) or infinite (weak ω-groupoids)
\cite{DBLP:conf/csl/AltenkirchR12}. %
%
Our work here concentrates on the internalization in \Coq of the setoid
model of type theory, seen as degenerate groupoids, where we can have a
self-contained definition of the structures involved. 
% Our long term goal
% is to formalize Hofmann and Streicher's groupoid model, which
% generalizes the construction from setoids to groupoids, but interresting
% difficulties already arise in this simpler setting.
%
Our first motivation to implement this translation is to explore the
interpretation of type theory in setoids in a completely intensional
setting and in the type theoretic language, leaving no space for
imprecision on the notions of equality and coherence involved.  We also
hope to give with this translation a basic exposition of the possible
type theoretic implications of the setoid/groupoid/homotopy models,
bridging a gap in the literature.
%
On the technical side, we have to slightly move away from existing
interpretations like Hofmann and Streicher's, for two important reasons
that we discuss now.


\paragraph{\bf A univalent model that makes no use of univalence.}


Our long term goal is to provide an interpretation of homotopy type
theory into type theory (without extensional principles).
%
This would give a meaning to all extensionality
principles without relying on them in the target theory.

However, if we use a traditional approach and formalize groupoid laws
using the identity type, it turns out that the type of isomorphisms
between two objects $x$ and $y$ of a groupoid, noted $x \sim_1 y$, must
also be formalized using the identity type. Or, putting it in the
homotopy type theoretic language, we have to consider univalent
groupoids.
%
This is because, in the groupoid interpretation, we need isomorphisms
between groupoids to satisfy groupoids laws using the identity
type as the notion of equality between isomorphisms. Then, using
functional extensionality, this amounts to compare the two functions
pointwise, but again using the identity type, and not the internal
notion of equality in the groupoid, given by $\sim_1$.
%
This means that $\sim_1$ has to coincide with the identity type, which
is precisely the property of being a univalent groupoid.
%
But then, this means that isomorphisms between groupoids should be
reflected in the identity type, which forces the target theory to
satisfy univalence already\ldots
%
So the first conclusion is that an internalization of the interpretation
in the style of formalization of categories presented in~\cite{hottbook}
would require the target theory to be univalent already.
 
To avoid this issue in our internalization of groupoids, the groupoid
laws are imposed using a notion of 2-dimensional equality that does
not have to be the identity type, just an equivalence relation.
%
Then, to enforce that the types of isomorphisms constitute
(homotopical) sets, we still use the identity type, but only to
express triviality of higher dimensions, not coherences themselves.
%
This interpretation of strictness is closer to the idea that a
groupoid is a weak ω-groupoid for which all equalities at dimension 2
are the same.
%
%
Note that our presentation requires less properties on identity types,
but we still need the axiom of functional extensionality to prove
triviality of higher dimensions for the groupoid of functors.
%
But this indicates that if we scale to ω-groupoids, the presence of
identity types in the core type theory will not be necessary anymore
and so the core type theory will be axiom free.
%
Thus, this paper can be seen as a proof of concept that it is possible
to interpret type theory into type theory without identity types.


\paragraph{\bf Taking dimension issues into account.}
%

In Hofmann and Streicher's groupoid model, a type $A$ depending on
context $\Gamma$ is interpreted as a functor 
$$
\sem{A} \ : \ \sem{\Gamma} \longrightarrow \GPD
$$
%
where $\GPD$ is said to be \emph{``the (large) category of groupoids''}. 
%
This means that they consider small groupoids, 
i.e., groupoids for which the type of objects is also a set, and
rely on set-theoretic extensional equality to witness natural
isomorphisms between groupoid isomorphisms. This is fine as long as we
consider set theory.
%
% But groupoids form naturally a 2-category/2-groupoid and not a
% category/groupoid, because functors/isomorphisms between (large)
% groupoids do not form a set but an arbitrary collection (a type in type
% theory). Otherwise said, groupoids are not enriched over themselves.
%
% Hofmann and Streicher have solved this issue as we usually do in
% category theory, by considering small groupoids, that is groupoids for
% which the type of objects is also a set, and relying on set-theoretic
% extensional equality to witness natural isomorphisms between groupoid 
% isomorphisms. This is fine as long as we consider set theory.
%
But when moving to type theory, small groupoids that form a category
actually correspond to setoids, unless we rely on axioms. Indeed, to
impose the smallness condition internally, we have to impose either that
(i) the identity type on the type of objects is an h-prop, or that (ii)
$x \sim_1 y$ is an h-prop for all objects $x$ and $y$. But condition (i)
gives rise to a category only in presence of functional extensionality,
which we precisely want to interpret.
%
This means that only condition~(ii) provides a correct notion
of small groupoids in type theory without using functional
extensionality. But condition~(ii) is the definition of a setoid.
%
% Note that this is also true for the notion of univalent groupoids
% introduced above as in that case, $\sim_1$ coincides with the identity type.
%
% In other words, using an internalization of groupoids, one can only
% interpret types as functors into the groupoid of setoids. This is what we do
% in this paper, which prevents us from giving a complete formulation of
% univalence in the source theory. That is, the invariance by
% isomorphism principle is respected in the model but has no counterpart
% in the source theory. 

Another way to solve this size issue would be to formalize the notion of
2-groupoids instead. However, the 1-groupoid case already raises
important challenges that wouldn't go away in the 2-groupoid case. 
Moreover, we believe that the real challenge is to interpret
$\omega$-groupoids which is the proper notion of self-enriched 
groupoid-like structure and is the subject of on-going work.

% coherences even limited at
% groupoids, the the structures are already quite tedious to manipulate and we
% found some interesting conditions on the structure of $Π$ and $Σ$ types
% that we believe were never presented in this form before. In the
% development, we strived to be as generic as possible and use the
% abstract structures of category theory to not be essentially tied to a
% particular dimension. That level of genericity also relies on the
% extension of the system with universe polymorphism.

% This translation could also be used to complete a forcing
% translation of type theory into type theory~\cite{hal-00685150}. By
% building up forcing layers on top of a core type theory, one can
% introduce new logical operators or type constructors (e.g. modal logic,
% recursive types), defined by translation. For correctness, this
% translation relies on a type theory that integrates proof-irrelevance
% and functional extensionality. The present translation gives the
% expressive power needed to compose with the forcing translation and get
% a fully definitional extension of type theory with forcing.

% To summarize, our contributions are: 
% \begin{enumerate}
% \item A theory of groupoids compatible with the ω-groupoids approach
%   and a partially mechanized interpretation of type theory into groupoids.
% \item A type-theoretic description of the necessary conditions on the
%   interpretation of function types, including dependent product types
%   which give rise to dependent groupoid functors and dependent sums.
% \item A proof of concept that it is possible to interpret homotopy
%   type theory into type theory without identity types.
% \end{enumerate}

\paragraph{\bf Outline of the paper.}
% The paper is organized as follows: 
Section \ref{sec:setting-translation} introduces the source type theory
of the translation and some features of the proof assistant that are
used in the formalization. The formal model includes a formalization of
groupoids and associated structures (\S\ref{sec:w2gpds}-\ref{sec:rew})
and a construction of the groupoids interpreting the standard type
constructors (\S\ref{sec:depprod}-\ref{sec:sigma}). Section
\ref{sec:interpretation} presents the internalization of the model
proper, and Section~\ref{section:cwf} shows that it constitutes (almost)
an internal category with families.
% 
 Section \ref{sec:conclusion} concludes on related and future work.
%

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 