\documentclass[runningheads,a4paper]{llncs}
\usepackage[T1]{fontenc}

\usepackage{utf}
\usepackage{float}
\usepackage{stfloats}
%\usepackage{midfloat}
%\usepackage{dblfloatfix}
\usepackage{subfig}
\usepackage{latexsym,amssymb,amsmath,ae,aeguill,amscd,stmaryrd}
\let\mathbb\undefined  % delete the command definition 
\usepackage{bbold}     % let bbold define its own \mathbb command
%\usepackage{mathrsfs}
\usepackage{xcolor}
\usepackage{tikz,xspace}
\usetikzlibrary{arrows,automata}
\usepackage[]{hyperref}

% My packages
\usepackage{abbrevs}
\usepackage{coq}
\usepackage{code}
\usepackage{qsymbols}

\usepackage[color]{coqdoc}
\definecolor{kwblack}{rgb}{0,0,0}
\def\coqdockwcolor{kwred}
\renewcommand{\coqdocnotation}[1]{#1}
\renewcommand{\coqexternalref}[3]{\href{#2.html\##1}{\textcolor{black}{#3}}}

\renewcommand{\coqdef}[3]{\phantomsection\hypertarget{coq:#1}{\text{#3}}}

\setlength{\coqdocbaseindent}{0.5em}

\newcommand{\aidememoire}[1]{\textcolor{red}{[[ #1 ]]}}
\def\interp#1{\ensuremath{\llbracket #1 \rrbracket}}

\newcommand{\mynote}[2]{
    \fbox{\bfseries\sffamily\scriptsize#1}
    {\small$\blacktriangleright$\textsf{\emph{#2}}$\blacktriangleleft$}~}

\newcommand\nt[1]{\mynote{\textcolor{magenta}{NT}}{#1}}
\newcommand\ms[1]{\mynote{\textcolor{blue}{MS}}{#1}}

\usepackage{url}

\input{univmacros}

\def\textchi{\ensuremath{\chi}}
\def\textbeta{\ensuremath{\beta}}
\DeclareUnicodeCharacter{8657}{\ensuremath{\uparrow}}

\def\coqlibrary#1#2#3{}

\renewenvironment{coqdoccode}{\begin{footnotesize}}{\end{footnotesize}}
\def\coqcode#1{#1}

% \renewcommand{\coqdocemptyline}{\vspace{0.3em}}

\def\kw#1{\mathsf{#1}}
\def\coqdoccst#1{{\color{\coqdoccstcolor}{\textsf{#1}}}}
\def\cst#1{{\color{\coqdoccstcolor}{#1}}}

\def\eqty#1#2#3{\kw{eq}~#1~#2~#3}
\def\matcheqkw{\kw{matcheq}}
\def\matcheq#1#2#3{\matcheqkw\ #1\ \kw{in}\ #2\ \kw{gives}\ #3}
\def\vec#1{\ensuremath{\overrightarrow{#1}}}
\def\nfdelta{{\downarrow^{δ}}}
\def\nfdeltab#1{#1{\downarrow^{δ}}}
\def\hnfbetadelta{\downarrow_{βδ}}
\def\lrule#1{\RefTirName{#1}}

%\renewcommand{\irule}[3]
%{\inferrule[#1] {\scriptstyle#2 }{\scriptstyle#3 }}

\begin{document}

\mainmatter  % start of an individual contribution

% \title{Internalization of the Groupoid Interpretation of Type Theory}
\title{A mechanized setoid model of Type Theory based on an
  internalization of groupoids}
\titlerunning {A mechanized model of Type Theory based on groupoids}

\author{Matthieu Sozeau\inst{1,2} \and Nicolas Tabareau\inst{1,3}}

\date{}

\institute{
$\pi r^2$ and Ascola teams, INRIA \and 
Preuves, Programmes et Systèmes (PPS) \and 
Laboratoire d'Informatique de Nantes Altantique (LINA)  
\\
\email{firstname.surname@inria.fr}
}


\def\mathrm#1{#1}

\maketitle
%  Already,formalizing such weak structures is a difficult endeavor. 

\begin{abstract}

  Homotopical interpretations of Martin-Löf type theory lead toward an
  interpretation of equality as a richer, more extensional
  notion. Extensional or axiomatic presentations of the theory with
  principles based on such models do not yet fully benefit from the
  power of dependent type theory, that is its computational
  character. Reconciling intensional type theory with this richer notion
  of equality requires to move to higher-dimensional structures where
  equality reasoning is explicit, which explains the emerging interest on
  simplicial or cubical models. 
  %
  However, such models are still based on set theory, which is somehow
  in opposition with the goal to replace set theory by homotopy type
  theory in the foundations of mathematics. 
  %
  Therefore, it is important to be able to express those models
  directly in type theory.
  % 
  In this paper, we follow this idea and develop an internalization of
  a setoid model of Martin-Löf type theory based on an internalization
  of groupoids.
% and respecting the invariance by isomorphism principle.
  % 
  Our work shows that even a simple setoid model of type theory does
  not form a model (at least in the sense of categories with families)
  when internalized inside type theory, as the distinction between
  definitional equality and equality in the model forbids some
  equalities on substitutions to hold without explicit rewriting.
  % 
  The observation is already present in the seminal work of Dybjer on
  internal categories with families. 
  % 
  Our formal development relies crucially on ad-hoc polymorphism to
  overload notions of equality and on a conservative extension of
  Coq's universe mechanism with polymorphism.
  % 

  % Homotopical interpretations of Martin-Löf type theory lead toward an
  % interpretation of equality as a richer, more extensional
  % notion. Extensional or axiomatic presentations of the theory with
  % principles based on such models do not yet fully benefit from the
  % power of dependent type theory, that is its computational
  % character. Reconciling intensional type theory with this richer notion
  % of equality requires to move to higher-dimensional structures where
  % equality reasoning is explicit. In this paper, we follow this idea and
  % develop an internalization of a model of Martin-Löf type theory based
  % on groupoids and respecting the invariance by isomorphism principle.
  % Our formal development relies crucially on ad-hoc polymorphism to
  % overload notions of equality and on a conservative extension of
  % Coq's universe mechanism with polymorphism.
\end{abstract}



\input{introduction}

\input{setting}

\section{Formalization of groupoids}
\label{sec:formalization}

\input{../theories/groupoid}

\section{The setoid interpretation}
\label{sec:interpretation}


\input{../theories/groupoid_interpretation_def}

\input{../theories/groupoid_interpretation}

\input{../theories/fun_ext}

\input{../theories/cwf_equations}


% \input{type-theory-elab}

%\input{bench-apps}

\input{related-concl}

% This is the text of the appendix, if you need one.

% \acks

% Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{splncs}
\bibliography{biblio}

% \newpage

% \appendix
% \section{Algebraic universes}
% \label{sec:algunivs}
% \input{appendix_univs}


% \newpage
% \appendix
% \section{Algebraic universes}
% \label{sec:algunivs}
% \input{appendix_univs}

% The bibliography should be embedded for final submission.

% \begin{thebibliography}{}
% \softraggedright

% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
% P. Q. Smith, and X. Y. Jones. ...reference text...

% \end{thebibliography}

\end{document}
